Increasing Chlorophyll *a* Amid Stable Nutrient Concentrations in Rhode Island Lakes and Reservoirs

### 2020-02-11 manuscript/peerj_preprint/manuscript.Rmd
Ready to submit pre-print

---
title: "Increasing Chlorophyll *a* Amid Stable Nutrient Concentrations in Rhode Island Lakes and Reservoirs"
author:
- Hollister. J. W. ^\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Kreakie, B. J. ^*1*^, Shivers,
  S. ^*3*^, Milstead, W. B. ^*1*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A.
  ^*2*^
- ^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
  Coastal Environmental Sciences Division, Narragansett, RI 02882
- ^*2*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
  RI 02881
- ^*3*^ ORISE, Narragansett, RI 02882
- '^\*^  *corresponding author: hollister.jeff@epa.gov*'
bibliography: manuscript.bib
output:
  word_document:
    reference_docx: manuscript_template.docx
editor_options:
  chunk_output_type: console
csl: ecology.csl
abstract: 'Addressing anthropogenic impacts on aquatic ecosystems is a focus of lake management. Controlling phosphorus and nitrogen can mitigate these impacts, but determining management effectiveness requires long-term datasets. Recent analysis of the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) United States found stable water quality in the northeastern and midwestern United States, however, sub-regional trends may be obscured. We analyze a sub-regional (i.e., 3000 km^2^) trend with the University of Rhode Island’s Watershed Watch Volunteer Monitoring Program (URIWW) dataset. URIWW has collected water quality data on Rhode Island lakes and reservoirs for over 25 years. The LAGOSNE and URIWW datasets allow for comparison of water quality trends at regional and sub-regional extents, respectively. We assess regional (LAGOSNE) and state (URIWW) trends with yearly averaged anomalies calculated on a per-station basis. Sub-regionally, temperature and chlorophyll *a* increased from 1993 to 2016. Total nitrogen shows a weak increase driven by low years in the early 1990s. Total phosphorus and the nitrogen:phosphorus ratio (N:P) were stable. At the regional scale, the LAGOSNE dataset shows similar trends to prior studies of the LAGOSNE with chlorophyll *a*, total nitrogen, total phosphorus, and N:P all stable over time. In short, algal biomass, as measured by chlorophyll *a* in Rhode Island lakes and reservoirs is increasing, in spite of stability in total nitrogen, total phosphorus, and the nitrogen to phosphorus ratio. This analysis suggests an association between lake temperature and primary production. Additionally, we demonstrate both the value of long-term monitoring programs, like URIWW, for identifying trends in environmental condition, and the utility of site-specific anomalies for analyzing for long-term water quality trends.'
---

@@

Examining long-term trends in Rhode Island lakes is possible because of the data gathered by University of Rhode Island’s Watershed Watch (URIWW).   URIWW is a scientist-led citizen science program founded in the late 1980s that has built a robust collaboration between URI scientists and a vast network of volunteer monitors.  Volunteer monitors are trained and then collect *in situ* data as well as whole water samples during the growing season (e.g., May through October).  The entire effort follow rigorous quality control/quality assurance protocols.  These types of citizen science efforts allow for the collection of reliable data that in turn lead to crucial and frequently unexpected insights [@dickinson2012current; @kosmala2016assessing; @Oliver_2017].  URIWW data contributed to the larger regional study by Oliver et al. [-@Oliver_2017], and, also allowed us to examine the long-term trends specifically in Rhode Island.

@@

The study area for this analysis includes lakes and reservoirs in the state of Rhode Island where data were collected by the University of Rhode Island's Watershed Watch program (`r fig_num("ww_map", display = "cite", caption = "Map of URI Watershed Watch lake and reservoir sampling sites")`).  The URIWW program began in 1988, monitoring 14 lakes and has now grown to include over 250 monitoring sites on over 120 waterbodies, including rivers/streams, and estuaries, with more than 400 trained volunteers. URIWW now provides more than 90% of Rhode Island's lake baseline data and is an integral part of the state's environmental data collection strategy. Data quality assurance and control is treated with paramount importance; volunteers are trained both in the classroom and the field, regular quality checks occur, and volunteers are provided with all the necessary equipment and supplies, along with scheduled collection dates. For freshwater lakes and reservoirs, weekly secchi depth and water temperature are recorded, along with bi-weekly chlorophyll *a* and in deep lakes (greater than 5 meters) dissolved oxygen. Water samples are collected three times per season (May through October) to be analyzed for nutrients.

@@

For this analysis, we were interested in trends in lake temperature, TN, TP, TN:TP, and chlorophyll *a*.  In particular, we selected URIWW data that matched the following criteria: 1) were sampled between 1993 and 2016, 2) were sampled in May to October, 3) and were sampled at a depth of 2 meters or less.  For each of these parameters, we further filtered the data to select sites that had at least 10 years of data for a given parameter within the 1993 to 2016 time frame.  The final dataset used in our analysis included `r nrow(ww_avail_data_locations)` lakes and reservoirs.  Of these sites, our filtered dataset had approximately `r length(temp_10yr_sites)` sites measured for temperature, `r length(chla_10yr_sites)` sites measured for chlorophyll *a*, `r length(total_n_10yr_sites)` sites measured for TN, and `r length(total_p_10yr_sites)` sites measured for TP.  Of the `r nrow(ww_avail_data_locations)` sampling sites, `r sum(ww_avail_data_locations$params_avail == "chla, total_p, total_n, np, temp")` had data for all 5 parameters.  The N:P ratio was calculated by dividing the mass concentrations of total nitrogen and total phosphorus and then converting to a molar ratio by multiplying by 2.21 (e.g., atomic weight of P 30.974/atomic weight of N 14.007).

@@

Methods for calculating the site-specific anomalies and the yearly averages are as follows and are presented graphically in `r fig_num(name = "anomaly_schematic", caption = "Example calculation of the site-specific anomalies and yearly mean anomalies.", display = "cite")`.  Additionally, an example R script, `schematic_anomaly.R` and example dataset, `schematic.csv` to recreate and demonstrate the calculations in `r fig_num("anomaly_schematic", display = "cite")` is available from "fill in with zenodo link" [@zenodo_citation].

@@

Testing for a regression slope being different than zero can be used to test for monotonic trends in water quality data [@helsel2002statistical].  We used these standard procedures to test for positive or negative trends in lake temperature, chlorophyll *a*, TN, TP and TN:TP.  For each parameter, we fit a regression line to the anomalies as a function of year and tested the null hypothesis that no trend existed (e.g., $\beta$~1~ = 0). The slope of this line provides information on the average yearly change of that paramter over the time period studied.  Traditionally, trends would be determined by assessing "significance" but recent guidelines suggest not using arbitrary p-value cut-offs to assesses significance [@wasserstein2016asa].  Our interpretation of the trends attempts to follow this advice and we assess trends with the information provided by the magnitude of the slopes, the p-values, and our understanding of the processes involved.

@@

Average annual temperature anomalies in lakes and reservoirs appears to be increasing (slope = `r signif(ww_temp_rl$slope,2)` , p = `r signif(ww_temp_rl$p.value, 2)`) with the majority of years with average temperature greater than the long-term average occurring in recent years (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature (average anomaly) in Rhode Island lakes and reservoirs. Points are averages of site-specific anomalies and ranges are standard deviations of site-specific anomalies. Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term means. Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term means.", display = "cite")`).  Chlorophyll *a* is also showing an increasing trend over time (slope = `r signif(ww_chla_rl$slope,2)` , p = `r signif(ww_chla_rl$p.value, 2)`) and with the exception of a slightly above-average year in 2003, the above-average years have all occurred in the most recent years (`r fig_num(name = "chla_trend", caption = "Yearly trend over 20+ years of chlorphyll a (average anomaly).  Panel A. Yearly averaged chlorophyll a anomalies from the URI Watershed Watch data.  Panel B. Yearly averaged chlorophyll a anomalies from the LAGOSNE dataset. Points are averages of site-specific anomalies and ranges are standard deviations of site-specific anomalies. Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term means. Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term means.", display = "cite")`A.).

@@

Mean annual trends for nutrients were weaker or showed no trend over time. The data suggest a positive trend in TN (slope = `r signif(ww_tn_rl$slope,2)` , p = `r signif(ww_tn_rl$p.value, 2)`); however, that perceived trend is driven by the lower than average TN values in 1993 and 1994 (`r fig_num(name = "tn_trend", caption = "Yearly trend over 20+ years of TN (average anomaly). Panel A. Yearly averaged TN anomalies from the URI Watershed Watch dataset.  Panel B. Yearly averaged TN anomalies from the LAGOSNE dataset. Points are averages of site-specific anomalies and ranges are standard deviations of site-specific anomalies. Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term means. Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term means.", display = "cite")`A.).  Since 1995, the yearly trend shows a lower increase over time (slope = `r signif(ww_tn_rl_no9394$slope,2)`, p = `r signif(ww_tn_rl_no9394$p.value, 2)`).  TP does not show a trend over time in the yearly anomalies (slope = `r signif(ww_tp_rl$slope,2)` , p = `r signif(ww_tp_rl$p.value, 2)`) and years that are over or under the average are more evenly distributed over the years (`r fig_num(name = "tp_trend", caption = "Yearly trend over 20+ years of TP (average anomaly).  Panel A. Yearly averaged TP anomalies from the URI Watershed Watch dataset.  Panel B. Yearly averaged TP anomalies from the LAGOSNE dataset. Points are averages of site-specific anomalies and ranges are standard deviations of site-specific anomalies. Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term means. Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term means.", display = "cite")`A.). The pattern is the same for the TN:TP ratio (slope = `r signif(ww_np_rl$slope,2)`, p = `r signif(ww_np_rl$p.value,2)`) with little evidence suggesting a change in the concentrations of TN relative to the concentrations of TP (`r fig_num(name = "np_trend", caption = "Yearly trend over 20+ years of the TN:TP ratio (average anomaly).  Panel A. Yearly averaged TN:TP ratio anomalies from the URI Watershed Watch dataset.  Panel B. Yearly averaged TN:TP ratio anomalies from the LAGOSNE dataset. Points are averages of site-specific anomalies and ranges are standard deviations of site-specific anomalies. Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term means. Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term means.", display = "cite")`A.).  Data for all figures are available as a comma-separated values file, `yearly_average_anomaly.csv` from "fill in with zenodo link" [@zenodo_citation].

@@

Our sub-regional analysis indicates that even when nutrient regimes exhibit relative stability (i.e., neither increasing nor decreasing over time), increases in primary production, as measured by chlorophyll *a*, can occur.  Over the same period we also demonstrate long-term warming of Rhode Island lakes and reservoirs. Chlorophyll has increased on average `r signif(ww_chla_rl$slope, 2)` μg/L per year over the 23 years of our analysis, while temperature has increased `r signif(ww_temp_rl$slope, 2)` °C per year over the same period.   This suggests that the observed increase in productivity, as measured by chlorophyll *a*, may be a result of warming waters and not a response to changes in nutrient condition.  Also, geographic extent does indeed matter when trying to identify long-term water quality trends. Similar to the results of Oliver et al. [-@Oliver_2017] our analysis shows little increasing trend in chlorophyll *a* at the regional scale (e.g., northeastern and mid-western United States).  However, at the local scale of the state of Rhode Island, there is a clear increasing trend in chlorophyll *a*.

@@

The analysis approach we used here, site-specific anomalies, is not a novel method and does have a long history in the analysis of trends in climate [@hansen2006global; @hansen2010global; @jones1996calculating; @jones1999surface]. However, using it to examine water quality trends is a new application of the technique, as we could find little evidence of using it specifically for water quality trends.  We built on these methods and adapted them for use with long-term water quality trends.  While other methods are valid and robust [e.g., @Oliver_2017], we chose averaged site-specific anomalies as they can provide readily interpretable results, especially for communicating to general audiences.  For instance, reporting the changes in anomalies allows us to look at changes in the original units.  With our analysis, the slope of the regression line for temperature suggests an average yearly increase of `r signif(ww_temp_rl$slope, 2)` °C and the slope of the regression line for chlorophyll *a* shows and average yearly increase of `r signif(ww_chla_rl$slope, 2)` µg/l.  Additionally, the site-specific anomalies are robust to variations in sampling effort and in the timing of inclusion of given sampling locations (e.g., added later in a time period or removed).  Lastly, this analysis is only possible because of the availability of sound, long-term data on water quality in Rhode Island.  Without the URIWW data and the commitment and participation of more than 2500 volunteers over the years, our analyses would have been impossible.  Going forward, it is important to appreciate the role that volunteer monitoring and citizen science programs can play in capturing and better understanding long term environmental trends.

>>>>

---
title: "Increasing Chlorophyll *a* Amid Stable Nutrient Concentrations in Rhode Island Lakes and Reservoirs"
author:
- Hollister. J. W. ^\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Kreakie, B. J. ^*1*^, Shivers,
  S. ^*3*^, Milstead, W. B. ^*1*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A.
  ^*2*^
- ^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
  Coastal Environmental Sciences Division, Narragansett, RI 02882
- ^*2*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
  RI 02881
- ^*3*^ ORISE, Narragansett, RI 02882
- '^\*^  *corresponding author: hollister.jeff@epa.gov*'
bibliography: manuscript.bib
output:
  word_document:
    reference_docx: manuscript_template.docx
editor_options:
  chunk_output_type: console
csl: ecology.csl
abstract: 'Addressing anthropogenic impacts on aquatic ecosystems is a focus of lake management. Controlling phosphorus and nitrogen can mitigate these impacts, but determining management effectiveness requires long-term datasets. Recent analysis of the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) United States found stable water quality in the northeastern and midwestern United States, however, sub-regional trends may be obscured. We analyze a sub-regional (i.e., 3000 km^2^) trend with the University of Rhode Island’s Watershed Watch Volunteer Monitoring Program (URIWW) dataset. URIWW has collected water quality data on Rhode Island lakes and reservoirs for over 25 years. The LAGOSNE and URIWW datasets allow for comparison of water quality trends at regional and sub-regional extents, respectively. We assess regional (LAGOSNE) and state (URIWW) trends with yearly mean anomalies calculated on a per-station basis. Sub-regionally, temperature and chlorophyll *a* increased from 1993 to 2016. Total nitrogen shows a weak increase driven by low years in the early 1990s. Total phosphorus and the nitrogen:phosphorus ratio (N:P) were stable. At the regional scale, the LAGOSNE dataset shows similar trends to prior studies of the LAGOSNE with chlorophyll *a*, total nitrogen, total phosphorus, and N:P all stable over time. In short, algal biomass, as measured by chlorophyll *a* in Rhode Island lakes and reservoirs is increasing, despite stability in total nitrogen, total phosphorus, and the nitrogen to phosphorus ratio. This analysis suggests an association between lake temperature and primary production. Additionally, we demonstrate both the value of long-term monitoring programs, like URIWW, for identifying trends in environmental condition, and the utility of site-specific anomalies for analyzing for long-term water quality trends.'
---

@@

Examining long-term trends in Rhode Island lakes is possible because of the data gathered by University of Rhode Island’s Watershed Watch (URIWW).   URIWW is a scientist-led citizen science program founded in the late 1980s that has built a robust collaboration between URI scientists and a vast network of volunteer monitors.  Volunteer monitors are trained and then collect *in situ* data as well as whole water samples during the growing season (e.g., May through October).  The entire effort follows rigorous quality control/quality assurance protocols.  These types of citizen science efforts allow for the collection of reliable data that in turn lead to crucial and frequently unexpected insights [@dickinson2012current; @kosmala2016assessing; @Oliver_2017].  URIWW data contributed to the larger regional study by Oliver et al. [-@Oliver_2017], and, also allowed us to examine the long-term trends specifically in Rhode Island.

@@

The study area for this analysis includes lakes and reservoirs in the state of Rhode Island where data were collected by the University of Rhode Island's Watershed Watch program (`r fig_num("ww_map", display = "cite", caption = "Map of URI Watershed Watch lake and reservoir sampling sites")`).  The URIWW program began in 1988, monitoring 14 lakes and has now grown to include over 250 monitoring sites on over 120 waterbodies, including rivers/streams, and estuaries, with more than 400 trained volunteers. URIWW now provides more than 90% of Rhode Island's lake baseline data and is an integral part of the state's environmental data collection strategy. Data quality assurance and control is treated with paramount importance; volunteers are trained both in the classroom and the field, regular quality checks occur, and volunteers are provided with all the necessary equipment and supplies, along with scheduled collection dates. For freshwater lakes and reservoirs, weekly secchi depth and water temperature are recorded, along with bi-weekly chlorophyll *a* and in deep lakes (greater than 5 meters) dissolved oxygen. Water samples are collected three times per season (May through October) to be analyzed for nutrients and bacteria.

@@

For this analysis, we were interested in trends in lake temperature, TN, TP, TN:TP, and chlorophyll *a*.  In particular, we selected URIWW data that matched the following criteria: 1) were sampled between 1993 and 2016, 2) were sampled in May to October, 3) and were sampled at a depth of 2 meters or less.  As not all sites have data for all selected years, we further filtered the data to select sites that had at least 10 years of data for a given parameter within the 1993 to 2016 time frame.  The final dataset used in our analysis included `r nrow(ww_avail_data_locations)` lakes and reservoirs.  Of these sites, our filtered dataset had approximately `r length(temp_10yr_sites)` sites measured for temperature, `r length(chla_10yr_sites)` sites measured for chlorophyll *a*, `r length(total_n_10yr_sites)` sites measured for TN, and `r length(total_p_10yr_sites)` sites measured for TP.  Of the `r nrow(ww_avail_data_locations)` sampling sites, `r sum(ww_avail_data_locations$params_avail == "chla, total_p, total_n, np, temp")` had data for all 5 parameters.  The N:P ratio was calculated by dividing the mass concentrations of total nitrogen and total phosphorus and then converting to a molar ratio by multiplying by 2.21 (e.g., atomic weight of P 30.974/atomic weight of N 14.007).

@@

Methods for calculating the site-specific anomalies and the yearly means are as follows and are presented graphically in `r fig_num(name = "anomaly_schematic", caption = "Example calculation of the site-specific anomalies and yearly mean anomalies.", display = "cite")`.  Additionally, an example R script, `schematic_anomaly.R` and example dataset, `schematic.csv` to recreate and demonstrate the calculations in `r fig_num("anomaly_schematic", display = "cite")` is available from "fill in with zenodo link" [@zenodo_citation].

@@

Testing for a regression slope being different than zero can be used to test for monotonic trends in water quality data [@helsel2002statistical].  We used these standard procedures to test for positive or negative trends in lake temperature, chlorophyll *a*, TN, TP and TN:TP.  For each parameter, we fit a regression line to the anomalies as a function of year and tested the null hypothesis that no trend existed (e.g., $\beta$~1~ = 0). The slope of this line provides information on the mean yearly change of that paramter over the time period studied.  Traditionally, trends would be determined by assessing "significance" but recent guidelines suggest not using arbitrary p-value cut-offs to assesses significance [@wasserstein2016asa].  Our interpretation of the trends attempts to follow this advice and we assess trends with the information provided by the magnitude of the slopes, the p-values, and our understanding of the processes involved.

@@

Mean annual temperature anomalies in lakes and reservoirs appears to be increasing (slope = `r signif(ww_temp_rl$slope,2)` , p = `r signif(ww_temp_rl$p.value, 2)`) with the majority of years with mean temperature greater than the long-term mean occurring in recent years (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature (mean anomaly) in Rhode Island lakes and reservoirs. Points are means of site-specific anomalies and ranges are standard deviations of site-specific anomalies. Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term means. Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term means.", display = "cite")`).  Chlorophyll *a* is also showing an increasing trend over time (slope = `r signif(ww_chla_rl$slope,2)` , p = `r signif(ww_chla_rl$p.value, 2)`) and with the exception of a slightly above-average year in 2003, the above-average years have all occurred in the most recent years (`r fig_num(name = "chla_trend", caption = "Yearly trend over 20+ years of chlorphyll a (mean anomaly).  Panel A. Yearly mean chlorophyll a anomalies from the URI Watershed Watch data.  Panel B. Yearly mean chlorophyll a anomalies from the LAGOSNE dataset. Points are means of site-specific anomalies and ranges are standard deviations of site-specific anomalies. Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term means. Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term means.", display = "cite")`A.).

@@

Mean annual trends for nutrients were weaker or showed no trend over time. The data suggest a positive trend in TN (slope = `r signif(ww_tn_rl$slope,2)` , p = `r signif(ww_tn_rl$p.value, 2)`); however, that perceived trend is driven by the lower than mean TN values in 1993 and 1994 (`r fig_num(name = "tn_trend", caption = "Yearly trend over 20+ years of TN (mean anomaly). Panel A. Yearly mean TN anomalies from the URI Watershed Watch dataset.  Panel B. Yearly mean TN anomalies from the LAGOSNE dataset. Points are means of site-specific anomalies and ranges are standard deviations of site-specific anomalies. Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term means. Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term means.", display = "cite")`A.).  Since 1995, the yearly trend shows a lower increase over time (slope = `r signif(ww_tn_rl_no9394$slope,2)`, p = `r signif(ww_tn_rl_no9394$p.value, 2)`).  TP does not show a trend over time in the yearly anomalies (slope = `r signif(ww_tp_rl$slope,2)` , p = `r signif(ww_tp_rl$p.value, 2)`) and years that are over or under the mean are more evenly distributed over the years (`r fig_num(name = "tp_trend", caption = "Yearly trend over 20+ years of TP (mean anomaly).  Panel A. Yearly mean TP anomalies from the URI Watershed Watch dataset.  Panel B. Yearly mean TP anomalies from the LAGOSNE dataset. Points are means of site-specific anomalies and ranges are standard deviations of site-specific anomalies. Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term means. Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term means.", display = "cite")`A.). The pattern is the same for the TN:TP ratio (slope = `r signif(ww_np_rl$slope,2)`, p = `r signif(ww_np_rl$p.value,2)`) with little evidence suggesting a change in the concentrations of TN relative to the concentrations of TP (`r fig_num(name = "np_trend", caption = "Yearly trend over 20+ years of the TN:TP ratio (mean anomaly).  Panel A. Yearly mean TN:TP ratio anomalies from the URI Watershed Watch dataset.  Panel B. Yearly mean TN:TP ratio anomalies from the LAGOSNE dataset. Points are means of site-specific anomalies and ranges are standard deviations of site-specific anomalies. Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term means. Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term means.", display = "cite")`A.).  Data for all figures are available as a comma-separated values file, `yearly_average_anomaly.csv` from "fill in with zenodo link" [@zenodo_citation].

@@

Our sub-regional analysis indicates that even when nutrient regimes exhibit relative stability (i.e., neither increasing nor decreasing over time), increases in primary production, as measured by chlorophyll *a*, can occur.  Over the same period we also demonstrate long-term warming of Rhode Island lakes and reservoirs. Chlorophyll has increased, on average, `r signif(ww_chla_rl$slope, 2)` μg/L per year over the 23 years of our analysis, while temperature has increased `r signif(ww_temp_rl$slope, 2)` °C per year over the same period.   This suggests that the observed increase in productivity, as measured by chlorophyll *a*, may be a result of warming waters and not a response to changes in nutrient condition.  Also, geographic extent does indeed matter when trying to identify long-term water quality trends. Similar to the results of Oliver et al. [-@Oliver_2017] our analysis shows little increasing trend in chlorophyll *a* at the regional scale (e.g., northeastern and mid-western United States).  However, at the local scale of the state of Rhode Island, there is a clear increasing trend in chlorophyll *a*.

@@

The analysis approach we used here, site-specific anomalies, is not a novel method and does have a long history in the analysis of trends in climate [@hansen2006global; @hansen2010global; @jones1996calculating; @jones1999surface]. However, using it to examine water quality trends is a new application of the technique, as we could find little evidence of using it specifically for water quality trends.  We built on these methods and adapted them for use with long-term water quality trends.  While other methods are valid and robust [e.g., @Oliver_2017], we chose mean site-specific anomalies as they can provide readily interpretable results, especially for communicating to general audiences.  For instance, reporting the changes in anomalies allows us to look at changes in the original units.  With our analysis, the slope of the regression line for temperature suggests a mean yearly increase of `r signif(ww_temp_rl$slope, 2)` °C and the slope of the regression line for chlorophyll *a* shows a mean yearly increase of `r signif(ww_chla_rl$slope, 2)` µg/l.  Additionally, the site-specific anomalies are robust to variations in sampling effort and in the timing of inclusion of given sampling locations (e.g., added later in a time period or removed).  Lastly, this analysis is only possible because of the availability of sound, long-term data on water quality in Rhode Island.  Without the URIWW data and the commitment and participation of more than 2500 volunteers over the years, our analyses would have been impossible.  Going forward, it is important to appreciate the role that volunteer monitoring and citizen science programs can play in capturing and better understanding long term environmental trends.
